---
title: "Simulated Annealing Schedules"
author: "Michael Hahsler"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Cooling Schedules

## Exponential Schedule
See [1]
```{r}
T_exp <- function(t, T0, alpha) T0 * alpha ^ t 
```

## Linear Schedule
See [1]
```{r}
T_linear <- function(t, T0, eta) T0 - eta * t 
```


## Fast Simulated Annealing
See [3]
```{r}
T_fast <- function(t, T0) T0 / (1 + t) 
```

## Logarithmic (Classic) Cooling

If $T_0$ is greater or equal to to the largest energy barrier in the problem, then
this schedule is guaranteed to find the global optimum in the limit of infinite time [2].
```{r}
T_log <- function(t, T0) T0 / log(1 + t) 
```

# Comparison

```{r}
t <- 0:100

schedules <- cbind(t = t, 
      Linear = T_linear(t, 1, .01),
      `Exponential (alpha = 0.8)` = T_exp(t, 1, .8),
      `Exponential (alpha = 0.9)` = T_exp(t, 1, .9),
      Fast = T_fast(t, 1),
      Logarithmic = T_log(t, 1)
      )

matplot(schedules[, -1], type = "l", lty = 1, ylab = "T(t)/T0", xlab = "t")
legend("topright", legend = colnames(schedules)[-1], col = 1:4, lty = 1, bty = "n")
```

# Literature

1. Kirkpatrick S, Gelatt C D Jr and Vecchi M P 1983 Optimization by simulated annealing Science 220 671-7
2. Hajek B 1988 Cooling schedules for optimal annealing Math. Oper. Res. 13 311–29
3. Szu H., Hardley R. Fast simulated annealing, June 1987, Physics Letters A 122(3-4):157-162 DOI:10.1016/0375-9601(87)90796-1
