---
title: "Solving a Maze using Reinforcement Learning"
author: "Michael Hahsler"
format:
    html: 
       embed-resources: true
editor: visual
---

## Introduction

This report demonstrates how value iteration (model-based RL)
and q-learning (model-free RL) solve a maze similar to the mazes we have used 
for tree search.

## Install and Load Package markovDP

Install the current development version of \[markovDP\](<https://mhahsler.r-universe.dev/markovDP>).

```{r}
if (!"markovDP" %in% rownames(installed.packages())) 
  install.packages('markovDP', repos = c(
    'https://mhahsler.r-universe.dev', 
    'https://cloud.r-project.org'))

library(markovDP)
```

## Read a Maze

Mazes are so called *qridworlds* where the agent moves around in a 2d environment composed out of tiles.  
The markovDP package already comes with the mazes we use in class.

```{r}
maze_dir <- system.file("mazes", package = "markovDP")
dir(maze_dir)

maze <- gw_read_maze(file.path(maze_dir, "L_maze.txt"))
maze
```

```{r}
#| fig-width: 8
#| fig-height: 8
gw_plot(maze)
```

The transition model is stored as a function.

```{r}
#| fig-width: 8
#| fig-height: 8
maze$transition_prob
gw_plot_transition_graph(maze)
```

The reward model

```{r}
maze$reward
```

The reward is a cost of 1 for each transition and a reward of +100 for reaching the goal state. The best solution will be the shortest path to the goal.

Remember that tree search starts with the start state and explores the space till it finds the goal state.

## Solve the Maze MDP as a Planning Agent

We solve the MDP directly using the problem description as a model for the environment. This approach would be used by a **planning agent** similar to the tree search algorithms.

Since the agent uses a complete model of the environment, we use a model-based reinforcement learning algorithm. Here we use value iteration. The algorithm sweeps in each iteration through all states and updates each state value using the utility of the best state it could get to plus the immediate reward. It stops when the state values do not change anymore (less than a set error threshold).

```{r}
#| fig-width: 8
#| fig-height: 8
system.time(sol <- solve_MDP(maze, method = "value_iteration"))

sol

gw_plot(sol)
gw_path(sol)
```

The color represents the state value. Here are the learned state values and the optimal action for each state (policy).

```{r}
policy(sol)
```

Solve the maze step-by-step.

```{r}
#| fig-width: 8
#| fig-height: 8
gw_animate(maze, "value_iteration", n = 25, zlim = c(-1,100))
```

Value iteration looks a lot like breath-first search starting at the goal or 
like a [flood fill](https://en.wikipedia.org/wiki/Flood_fill) algorithm. 
The goal is to find the **optimal state value for each state.**

The space complexity is bad with $O(|S|)$ and the number of states is typically
very large.


## Reinforcement Learning for Unknown Mazes

An unknown environment mean that we do not have a model of the environment. 
The agent does not know the transition function or the reward structure and has 
to **learn a good policy** and at least part of the transition and reward models.

Here we assume that the agent has no direct access to the MDP. 
The agent can only try actions and the environment uses the MDP to simulate how it reacts.
These methods are also called *model-free* approach for reinforcement learning.

### Monte Carlo Control

On-policy Monte Carlo control uses an exploring $\epsilon$-greedy policy to explore
the maze. It is called on-policy because it also learns the same $\epsilon$-greedy policy.

Monte Carlo methods creates on episode (search from the start to the goal) at a time and
then updates the state values in reverse order using the learning parameter $\alpha$.
This is how such an episode looks like.
```{r}
set.seed(1111)
sample_MDP(maze, n = 1, horizon = 50, trajectories = TRUE)$trajectories
```

After 50 steps, the random walk cannot find the goal, which is a problem. This is why 
I use a much larger horizon.

```{r}
system.time(sol <- solve_MDP(maze, method = "MC_on_policy", 
                 horizon = 1000, n = 100, epsilon =.2, alpha = 0.3))

```

```{r}
#| fig-width: 8
#| fig-height: 8
gw_plot(sol)
```

We see that it explors a large part of the maze, but then concentrates on the direct path 
to the goal.

Step-by-step solution.


```{r}
#| fig-width: 8
#| fig-height: 8
sol <- gw_animate(maze, "MC_on_policy", n = 25, 
                  horizon = 1000, epsilon = 0.2, alpha = 0.3, first_visit = FALSE)
```

Initially, the random exploration does not lead to the goal during the horizon, 
but eventually it finds a path. The behavior can be compared to 

* repeated depth-first search with a depth limited,
* or Monte Carlo tree search for game trees where the utility
  is propagated from the end of the game to the current state.


### Q-Learning

Q-learning is a popular model-free RL method that does not update with complete 
episodes, but it updates only fore one step at a time. 

```{r}
system.time(sol <- solve_MDP(maze, method ="q_learning", 
                             horizon = 1000, n = 100, alpha = 0.3))

sol
```


```{r}
#| fig-width: 8
#| fig-height: 8
gw_plot(sol)
```

Step-by-step Solution using Q-Learning.
Q-learning follows an $\epsilon$-greedy policy where it takes the current beat action, but
with a probability of $\epsilon$ uses a random action to perform exploration. Initially,
the agent has no good policy and essentially performs random walk.

I use 1000 as the horizon with the hope that it will randomly run into the goal.
After some experimentation, I settled on an $\epsilon$ of $0.2$ which means every 
5th action is chosen randomly and I use a
relatively high, fixed learning rate $\alpha$ of $0.3$

```{r}
#| fig-width: 8
#| fig-height: 8
sol <- gw_animate(maze, "q_learning", n = 25, zlim = c(-1,100), 
                  horizon = 1000, epsilon = 0.2, alpha = 0.3)
```

Model-free RL methods are typically horribly data inefficient, but they do 
not need a model of the environment and do not need to calculate state values for 
all states.

